---
title: "GDP and Education data"
author: "Gerardo Garza"
date: "March 21, 2017"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This document delineates the analysis performed on the data sets for the gross domestic product ranking table and the education statistics. Both provided by the world bank. 
The GDP data was downloaded from : https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FGDP.csv
The Education data was obtained from: https://d396qusza40orc.cloudfront.net/getdata%2Fdata%2FEDSTATS_Country.csv

The original data sources for the GDP and Education data are located respectively at:
http://data.worldbank.org/data-catalog/GDP-ranking-table
and
http://data.worldbank.org/data-catalog/ed-stats

### Project Structure
The project contains 2 folders; Data and Analysis.  
The Data folder contains the code to download ("download.R") and to clean and merge the data ("cleanMerge.R").    
The Analysis folder contains both "Analysis.R" and "Plots.R" which are used to do all of the calculations and analysis as well as create the plots for this project.
```{R}
source("Data/downloadData.R")
source("Data/cleanMerge.R")
source("Analysis/Analysis.R")
source("Analysis/Plots.R")
```

The messages listed above should indicate the data was downloaded with a SHA-1 hash which we use to make sure the data is consistent for reproducibility purposes.
Please note all of the NA's introduced by coercion were accounted for and cleaned during the analysis so disregard the Warning message.

## The Data
The data was initially downloaded and stored in 2 separate data frames "GDP.Data" for the GDP Data and "Education.Data" for the Education data.    
After the data was cleaned and analyzed it was then stored in the data frames "newGDP" and "newEducation".

The initial data frames "GDP.Data" and "Education.Data" were not modified as to maintain the original data from the download.

GDP.Data contains 330 records and 10 variables
```{R}
names(GDP.Data)
```

Education.Data contains 234 records and 31 variables
```{r}
names(Education.Data)
```


## Data Merge
To complete this analysis a merge between the Education data and the GDP data was necessary.   
After the data from "GDP.Data" and "Education.Data" were cleaned they were merged and a new data frame was created called "mergedEducationGDP".
The merge was done by using CountryCode as the binding variable.     
The resulting records after the merge totaled 190 observations but one of the variables included an NA  value so it was removed, leaving us with a total of 189 as we can see below.
```{R Merge}
dim(mergedEducationGDP)
```

After the merge only the variables that are pertaining to this analysis were kept.
The new data frame "mergedEducationGDP" contains the following:
```{r}
names(mergedEducationGDP)
```

## Countries with a low Gross Domestic Product
The top 15 countries with the worse GDP are listed as follows
```{R}
head(subset(mergedEducationGDP, select=c(CountryCode, Ranking, Country)),n=15)
```
Based on this observation we see St. Kitts and Nevis is the country with the 13th worse GDP.

##Average GDP rankings for High Income groups
The average GDP ranking for all high income groups (both OECD and nonOECD) is:
```{R}
mean(highIncome$Ranking)
```
The average ranking for only the "High income: OECD" is:
```{r}
mean(OECD$Ranking)
```
And the average ranking for only the "High income: nonOECD" is
```{r}
mean(noOECD$Ranking)
```

This analysis shows there is a big difference in GDC rankings between these 2 groups.     
Even though they are both on the "High income" group, the nonOECD group has a much lower mean ranking (greater GDP).

## Distribution of GDP for all countries
The distribution of GDP for all countries based on their Income Group is as follows:
```{R Plots}
PlotAllCountries
```

This plot shows the overall distribution of GDP in all the countries. There are only 3 above the 5,000,000 B in GDP mark and most of the countries GDP is too low in comparison to be representative on this plot.


### GDP distribution for all countries in their respective income group
```{R}
PlotAllCountriesbyIGroup
```

The results on this graph show the GDP distribution for the High income:nonOECD group is actually closer to the Low income group than it is to the other High income group (OECD) which is interesting. Also, the highest country in the nonOECD group does not go beyond 720,000B in GDP. The only other group with that low distribution is the Low income group.   
Another interesting find is that the country with the highest GDP in the Lower middle income group is actually higher than all but one (HIgher income OECD) of the countries in the other groups.

## Summary of statistics of GDP by income groups
```{R summaryStats}
tapply(mergedEducationGDP$GDP, mergedEducationGDP$`Income Group`, summary)
tapply(mergedEducationGDP$GDP, mergedEducationGDP$`Income Group`, sd)
```

Based on these results we can see that the distribution on each group is not an evenly distribution. The less dramatic one being the Low income group. The differences between the mininum and maximum values are really big and just by looking at the comparisson between the Maximum values to the rest of the quantiles we can suspect there are big outliers on this data.     
This is even more apparent on the High income: OECD and the Lower middle income group.


## GDP rankings divided into 5 quantile groups compared to each Income group.
```{R Table}
tablebyRankGroup
```

In this analysis we can see as expected there are no Low income countries in group 1 with the highest GDP but it is also interesting to see that the Lower middle income group has 5 countries that are among the group of nations with the highest rank, which is 1 more than the High income:nonOECD group has.

## Conclusion
The results from this analysis show that the distribution of GDP amoung the countries is very different, even for the countries with the same Income group. At this point we don't have the information to show how the groups were selected but based on the findings here, we see that there is very little if any correlation between the income group and GDP. In other words a higher income may not be equal to a higher GDC value.
WIth these results it is hard to do a fair comparisson between countries, even on the same group. There are one or two countries on each group with a GDP value that is much higher than the rest of the other countries and it makes it harder to compare between them. In order to do a more fair and accurate comparisson we would most likely need to remove those countries with the higher values.